A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-11-27 01_Introduction.Rmd
and more

Hardings narrative is, even when taking into account it's enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology: Instead of directly querrying quantitative data the typical process for the creation of large scale overviews is to excerpt the literature of all relevant local research traditions. The present article in contrast explores whether openly available bulk radiocarbon data could allow to reconstruct the trends within the distribution of burial rites. This could set the precedent for for a more reproducible and statistically sound reconstruction of the main shifts. The quantitative comparison of macroregional trends also opens the perspective on long term patterns of interaction intensity on a continental scale.

The spread of ideas in human populations over space and time is a major topic of *Cultural evolution theory* [@creanzaCulturalEvolutionaryTheory2017]. This framework is as least as old as theories of biological evolution and underwent a long and versatile research history from *Evolutionism* and *Social Darwinism* in the 19th and early 20th century, against *Cultural relativism* and back to *Neoevolutionism* up until the nineteen sixties and seventies [@petermannGeschichteEthnologie2004]. In the late seventies the the discussion took another turn away from social structure to the more general observation of human behaviour. Three major theoretical movements emerged in this context: *Evolutionary psychology*, *Human behavioural ecology* and *Dual inheritance theory* [@smithThreeStylesEvolutionary2000]. The latter -- heavily popularized by Richard Dawkin's *Memetics* [@dawkinsSelfishGene1976] -- establishes the notion of culture acting as secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007]. Just like in biological evolution there might be selection and mutation within *Social learning* and therefore methods of *Population genetics* could be employed within cultural anthropology -- of course with the much faster pace of cultural exchange in mind. A lot of theoretical and quantitative research arose from this assumptions: e.g. @cavalli-sforzaCulturalTransmissionEvolution1981, @boydCultureEvolutionaryProcess1985, @teltserEvolutionaryArchaeologyMethodological1995 or @obrienEvolutionaryArchaeologyTheory1996.

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase well-being and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work here are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

>>>>

Hardings narrative is, even when taking into account it's enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology: Instead of directly querying quantitative data the typical process for the creation of large scale overviews is to excerpt the literature of all relevant local research traditions. The present article in contrast explores whether openly available bulk radiocarbon data could allow to reconstruct the trends within the distribution of burial rites. This could set the precedent for for a more reproducible and statistically sound observation of the main shifts. The quantitative comparison of macroregional trends also paves the way for an analysis of patterns of interaction intensity on a continental scale.

Already the assumption that the cremation aspect of Early Bronze Age burial rituals in Britain can be directly compared to the same aspect in Hungary requires a theoretical foundation. *Cultural transmission theory* provides a framework where this claim can be explored as a hypothesis for further research. The spread of ideas in human populations over space and time is a major topic of *Cultural evolution theory* [@creanzaCulturalEvolutionaryTheory2017]. This line of thought is as least as old as theories of biological evolution and underwent a long and versatile research history from *Evolutionism* and *Social Darwinism* in the 19th and early 20th century, against *Cultural relativism* and back to *Neoevolutionism* up until the nineteen sixties and seventies [@petermannGeschichteEthnologie2004]. In the late seventies the the discussion took another turn away from social structure to the more general observation of human behaviour. Three major theoretical movements emerged in this context: *Evolutionary psychology*, *Human behavioural ecology* and *Dual inheritance theory* [@smithThreeStylesEvolutionary2000]. The latter -- heavily popularized by Richard Dawkin's *Memetics* [@dawkinsSelfishGene1976] -- establishes the notion of culture acting as secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007]. Just like in biological evolution there might be selection and mutation within *Social learning* and therefore methods of *Population genetics* could be employed within cultural anthropology -- of course with the much faster pace of cultural exchange in mind. A lot of theoretical and quantitative research arose from this assumptions: e.g. @cavalli-sforzaCulturalTransmissionEvolution1981, @boydCultureEvolutionaryProcess1985, @teltserEvolutionaryArchaeologyMethodological1995 or @obrienEvolutionaryArchaeologyTheory1996.

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase well-being and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite (or because?) their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].
